---
title: "multiRegFlow: scope, workflow and package map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multiRegFlow: scope, workflow and package map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(multiRegFlow)
```

## Purpose

`multiRegFlow` is a guided workflow for multiple regression in agricultural,
soil and plant sciences. It does not replace `lm()`, `glmnet`, `quantreg`,
`robustbase`, `mgcv`, `np`, or other specialized packages. Its purpose is to
make the statistical path explicit: define the scientific goal, fit a transparent
reference model, diagnose the specification, investigate multicollinearity and
influence, quantify uncertainty, explore selection or shrinkage when justified,
compare flexible alternatives, validate prediction out of sample, and record the
reasoning in an auditable report.

The central distinction is between **explanation/inference** and **prediction**.
An explanatory analysis asks how a response changes conditionally on the terms
in a scientific model. A predictive analysis asks how accurately new observations
can be predicted. Selection criteria, uncertainty statements and performance
metrics must be interpreted differently under these goals.

## Available modules

```{r}
mr_capabilities()
```

The capability table also provides a quick check of optional backends installed
in the current R library.

```{r}
subset(mr_capabilities(), available)
```

A third useful view is the set of modules that need optional packages.

```{r}
subset(mr_capabilities(), !is.na(package) & package != "")
```

## Teaching datasets

The package ships frozen **synthetic** datasets. They were designed to create
known teaching situations and must not be represented as observations from real
field trials.

```{r}
soil <- mr_example_data("soil_fertility")
plant <- mr_example_data("plant_growth")
coll <- mr_example_data("agronomy_collinear")
str(soil)
```

```{r}
nl <- mr_example_data("irrigation_nonlinear")
het <- mr_example_data("agronomy_heteroskedastic")
c(n_soil = nrow(soil), n_plant = nrow(plant), n_nonlinear = nrow(nl))
```

```{r}
summary(coll[c("N_rate_kg_ha", "N_uptake_kg_ha", "canopy_N_g_m2")])
```

The five datasets emphasize different questions:

| Dataset | Main teaching purpose |
|---|---|
| `soil_fertility` | explanatory OLS, effects, importance and reporting |
| `plant_growth` | physiology and biomass regression |
| `agronomy_collinear` | redundant N-status indicators and regularization |
| `irrigation_nonlinear` | functional-form failure and flexible alternatives |
| `agronomy_heteroskedastic` | robust covariance and wild bootstrap |

## A minimal explanatory workflow

```{r}
fit_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    exchangeable_K_mg_dm3 + soil_moisture_pct + season_rain_mm,
  data = soil,
  goal = "explanation"
)
fit_soil
```

```{r}
mr_diagnose(fit_soil)$fit
mr_collinearity(fit_soil)$VIF
mr_recommend(fit_soil)
```

The output is deliberately modular. A single omnibus table can hide the fact
that model fit, coefficient uncertainty, collinearity and predictive performance
answer different questions.

## A minimal predictive workflow

```{r}
fit_pred <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  data = plant,
  goal = "prediction"
)
mr_validate(fit_pred, v = 5, repeats = 3, seed = 2026)$summary
```

Out-of-fold predictions are generated without using an observation in the fit
that predicts that same observation. This is preferable to reporting training
`R-squared` as predictive evidence.

## Flexible alternatives are sensitivity analyses, not automatic upgrades

```{r}
fit_nl <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C,
  data = nl,
  goal = "prediction"
)
mr_diagnose(fit_nl)$specification
```

A significant functional-form screen is a reason to inspect the residual
pattern and agronomic mechanism. It is not proof that a particular nonlinear
model is correct. The dedicated flexible-alternatives vignette demonstrates GAM,
kernel, robust and quantile approaches.

## Recommended reading order

1. `v01-foundations-to-advanced-tutorial.Rmd`
2. `v02-diagnostics-influence.Rmd`
3. `v03-collinearity-vif-condition.Rmd`
4. `v04-selection-stability.Rmd`
5. `v05-regularization.Rmd`
6. `v06-bootstrap-resampling.Rmd`
7. `v07-importance-effects-shap.Rmd`
8. `v08-flexible-alternatives.Rmd`
9. `v09-validation-comparison.Rmd`
10. `v10-soils-case-study.Rmd`
11. `v11-plant-science-case-study.Rmd`
12. `v12-reporting-audit.Rmd`

The tutorials intentionally repeat the **workflow logic**, but avoid repeating
whole sections of methodological discussion. Each vignette has a distinct
teaching responsibility.
