---
title: "Foundations to advanced multiple regression in agricultural sciences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Foundations to advanced multiple regression in agricultural sciences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.8)
library(multiRegFlow)
```

## 1. Begin with the scientific question

Multiple regression is most useful when the model is connected to a scientific
question. Consider a synthetic soil-fertility dataset in which crop yield is
related to soil reaction, organic matter, texture, available nutrients, soil
water and seasonal rainfall.

```{r}
soil <- mr_example_data("soil_fertility")
head(soil)
```

If the objective is to estimate conditional associations and quantify
uncertainty, declare an explanatory goal.

```{r}
fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  data = soil,
  goal = "explanation"
)
summary(fit)
```

A slope is interpreted conditionally on the remaining terms. For example, the
coefficient of available P describes the expected change in yield per unit of P
when the other predictors in the model are held fixed. This is not automatically
a causal effect. Causal interpretation requires design and assumptions that go
beyond a regression coefficient.

## 2. Fit is not adequacy

`R-squared` describes variation explained in the observed sample. It does not
establish linearity, constant residual variance, absence of influential points,
or transportability to future observations.

```{r}
dg <- mr_diagnose(fit)
dg$fit
dg$rank
```

The package also reports a Breusch-Pagan screen and a fitted-value-power
specification screen.

```{r}
dg$heteroscedasticity
dg$specification
```

These are screening tools. A small p-value does not dictate a unique repair.
The next step should depend on the observed residual pattern and the agronomic
mechanism that could create it.

## 3. Inspect residuals and influence

```{r fig1, fig.cap="Standard OLS diagnostic plots for the soil-fertility reference model."}
par(mfrow = c(2, 2))
plot(fit$model)
par(mfrow = c(1, 1))
```

Observation-level influence diagnostics can be ranked without automatically
deleting any observation.

```{r}
infl <- mr_influence(fit)
head(infl[order(-infl$cooks_distance), ], 8)
```

The flags correspond to transparent screening cutoffs for leverage, Cook's
distance, DFFITS and DFBETA. A flagged observation can be scientifically
important precisely because it represents a rare soil-climate combination.
Sensitivity analyses should therefore compare conclusions with and without the
observation while preserving the original analysis unless a documented data
error or design-based exclusion exists.

## 4. Multicollinearity is about information

Agronomic predictors frequently describe overlapping processes. The dedicated
collinear dataset contains N rate, N uptake and canopy N, which are biologically
related.

```{r}
coll <- mr_example_data("agronomy_collinear")
fit_coll <- mr_fit(yield_t_ha ~ ., data = coll)
co <- mr_collinearity(fit_coll)
co$VIF
```

```{r}
co$condition_index
head(co$correlations, 6)
```

VIF quantifies how much coefficient variance is inflated by linear dependence
among predictors. Condition indices and variance-decomposition proportions help
identify multivariable dependency patterns. The package does not use a rule such
as "VIF > 10, therefore delete the variable". The scientific role of each
predictor matters.

## 5. Classical and robust covariance inference

When the conditional mean model is defensible but residual variance is not
constant, robust covariance estimation can protect coefficient inference without
changing the fitted mean.

```{r}
het <- mr_example_data("agronomy_heteroskedastic")
fit_het <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  data = het
)
mr_diagnose(fit_het)$heteroscedasticity
```

```{r}
classical <- mr_inference(fit_het, vcov = "classical")
hc3 <- mr_inference(fit_het, vcov = "HC3")
merge(classical[, c("term", "std_error")],
      hc3[, c("term", "std_error")], by = "term",
      suffixes = c("_classical", "_HC3"))
```

Cluster-robust covariance is useful when observations share a known grouping
unit and the independence assumption applies at the cluster rather than the
individual-observation level.

```{r}
het$block <- rep(seq_len(18), length.out = nrow(het))
fit_block <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  data = het
)
mr_inference(fit_block, vcov = "cluster", cluster = "block")
```

The synthetic `block` variable here is only a teaching device. In a real
experiment, clustering must follow the actual sampling or experimental design.

## 6. Selection is exploratory unless justified otherwise

AIC and BIC can help compare candidate specifications.

```{r}
aic_sel <- mr_select(fit, method = "aic")
bic_sel <- mr_select(fit, method = "bic")
aic_sel$selected
bic_sel$selected
```

Scientifically required covariates can be retained during stepwise exploration.

```{r}
keep_sel <- mr_select(fit, method = "aic",
                      keep = c("available_P_mg_dm3", "season_rain_mm"))
keep_sel$selected
```

For confirmatory inference, the uncertainty generated by searching many models
is not captured by the ordinary p-values in the selected model. Prespecification,
validation and sensitivity analyses are preferable to treating a data-driven
stepwise model as if it had been fixed in advance.

## 7. Best subsets and stability

```{r}
if (requireNamespace("leaps", quietly = TRUE)) {
  bs <- mr_select(fit, method = "best_subset")
  bs$variables_by_bic
  bs$variables_by_adjr2
}
```

Stability selection uses repeated subsampling with a selection procedure and is
especially valuable when the question is whether a predictor is repeatedly
selected rather than merely selected once.

```{r, eval=FALSE}
st <- mr_select(fit_coll, method = "stability", cutoff = 0.75, PFER = 1)
st$selected
st$selection_probabilities
```

## 8. Regularization for correlated predictors

Ridge shrinks correlated coefficients toward zero without forcing them to zero.
Lasso can set coefficients exactly to zero. Elastic net combines both penalties.

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  ridge <- mr_regularize(fit_coll, "ridge", nfolds = 5, seed = 100)
  lasso <- mr_regularize(fit_coll, "lasso", nfolds = 5, seed = 100)
  enet <- mr_regularize(fit_coll, "elastic_net", alpha = 0.5,
                        nfolds = 5, seed = 100)
  ridge$lambda_1se
  lasso$selected
  enet$selected
}
```

```{r regplot, fig.cap="Cross-validation curve for lasso in the collinear agronomic example."}
if (exists("lasso")) plot(lasso$model)
```

For highly correlated measurements of the same biological process, ridge is
often useful as a coefficient-stabilization sensitivity analysis. Lasso can be
unstable with strongly correlated predictors because it may choose one member
of a group and ignore another with similar information.

## 9. Bootstrap uncertainty

Pairs bootstrap resamples complete observational units and naturally carries the
observed predictor distribution into each replicate.

```{r}
boot_pairs <- mr_bootstrap(fit, R = 200, type = "pairs", seed = 7)
boot_pairs$intervals
```

Residual bootstrap holds the design matrix fixed and resamples residuals.

```{r}
boot_res <- mr_bootstrap(fit, R = 200, type = "residual", seed = 7)
boot_res$sign_stability
```

Wild bootstrap is useful when the residual variance changes with the predictors.

```{r}
boot_wild <- mr_bootstrap(fit_het, R = 200, type = "wild", seed = 7)
boot_wild$intervals
```

For publication-quality inference, use more replicates than the tutorial uses
for speed and examine Monte Carlo stability.

## 10. Different definitions of importance

A standardized slope compares predictors on a standard-deviation scale.
Partial R-squared measures the proportion of residual variance attributable to
a coefficient conditional on the other terms. Semi-partial R-squared measures
the unique contribution of a term to total response variation.

```{r}
mr_importance(fit, "standardized")
mr_importance(fit, "partial_r2")
mr_importance(fit, "semi_partial_r2")
```

LMG importance decomposes model R-squared across predictor orderings.

```{r}
if (requireNamespace("relaimpo", quietly = TRUE)) {
  lmg <- mr_importance(fit, "lmg")
  lmg$lmg
}
```

SHAP provides an additive contribution scale useful when comparing explanatory
patterns across model classes. It should not be re-labelled as a causal effect.

```{r}
if (requireNamespace("kernelshap", quietly = TRUE)) {
  sh <- mr_importance(fit, "shap")
  head(sh$S)
}
```

## 11. Marginal effects

Marginal effects are especially useful when the fitted specification contains
interactions or nonlinear terms.

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "available_P_mg_dm3")
}
```

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "season_rain_mm")
}
```

## 12. When the linear functional form is inadequate

The irrigation dataset contains nonlinear response surfaces.

```{r}
nl <- mr_example_data("irrigation_nonlinear")
lin_nl <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C +
    plant_stand_plants_m2,
  data = nl,
  goal = "prediction"
)
mr_diagnose(lin_nl)$specification
```

A GAM can preserve a structured additive interpretation while allowing smooth
responses.

```{r}
gam_nl <- mr_fit(
  yield_t_ha ~ s(irrigation_mm) + s(N_rate_kg_ha) + s(mean_temp_C),
  data = nl,
  engine = "gam",
  goal = "prediction"
)
summary(gam_nl)
```

Kernel regression is more flexible but can become difficult to interpret in
higher dimensions. Its bandwidth selection can also be computationally costly.

```{r, eval=FALSE}
ker_nl <- mr_kernel(lin_nl)
ker_nl$bandwidth
```

## 13. Quantile regression for low- and high-response conditions

```{r}
if (requireNamespace("quantreg", quietly = TRUE)) {
  qfit <- mr_quantile(fit_het, tau = c(0.25, 0.5, 0.75))
  summary(qfit$model)
}
```

A predictor can have a different association in low-yield environments than in
high-yield environments. Quantile regression estimates these conditional
relationships without reducing the response to its conditional mean.

## 14. Prediction requires out-of-sample evidence

```{r}
fit_pred <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    exchangeable_K_mg_dm3 + soil_moisture_pct + season_rain_mm,
  data = soil,
  goal = "prediction"
)
cv <- mr_validate(fit_pred, v = 5, repeats = 3, seed = 42)
cv$summary
```

Calibration slope near one means that the spread of predictions is reasonably
aligned with the spread of observed responses. Calibration intercept near zero
means that average predictions are not substantially shifted.

## 15. Compare models without mixing criteria

```{r}
small <- mr_fit(
  yield_t_ha ~ organic_matter_pct + available_P_mg_dm3 + season_rain_mm,
  soil, goal = "prediction"
)
cmp <- mr_compare(full = fit_pred, reduced = small,
                  v = 5, repeats = 3, seed = 42)
cmp$inference
cmp$prediction
```

A model may have a better AIC and still fail to improve predictive RMSE. The
package keeps these two evidence streams separate.

## 16. An integrated recommendation

```{r}
mr_recommend(fit_coll)
mr_recommend(fit_het)
mr_recommend(lin_nl)
```

The recommendation engine is intentionally conservative. It identifies signals
and reasonable sensitivity analyses but does not choose a scientifically correct
model on behalf of the analyst.

## 17. Reproducible reporting

```{r}
report_text <- mr_report(fit)
cat(substr(report_text, 1, 1200))
```

For a predictive analysis, validation can be included automatically.

```{r}
report_pred <- mr_report(fit_pred, validation = TRUE,
                         v = 5, repeats = 2, seed = 42)
cat(substr(report_pred, 1, 500))
```

## 18. Practical decision sequence

A defensible workflow for many agronomic multiple-regression problems is:

1. define the observational or experimental unit;
2. define the response and scientifically motivated predictors;
3. state whether the principal goal is explanation or prediction;
4. fit the OLS reference model;
5. inspect functional form, residual variance and influential observations;
6. evaluate collinearity as an information problem, not as a mechanical deletion rule;
7. use robust covariance or bootstrap when the mean specification is adequate but variance assumptions are doubtful;
8. treat automated selection as exploratory unless stronger methodology is used;
9. use ridge, lasso or elastic net when shrinkage is scientifically useful;
10. compare GAM, kernel, robust or quantile models only when the scientific question motivates them;
11. validate predictive claims out of sample;
12. report effect estimates, uncertainty, sensitivity analyses and limitations together.

This sequence is the organizing principle of the remaining vignettes.
