---
title: "Diagnostics, heteroscedasticity and influential observations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics, heteroscedasticity and influential observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.7)
library(multiRegFlow)
```

## Why diagnostics come before selection

A regression coefficient is meaningful only relative to the fitted
specification and the observations that support it. Diagnostics therefore come
before automated selection. This vignette separates four issues: functional
form, residual variance, unusual response values and unusual predictor
combinations.

## Example 1: a well-behaved soil model

```{r}
soil <- mr_example_data("soil_fertility")
fit_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    exchangeable_K_mg_dm3 + soil_moisture_pct + season_rain_mm,
  soil
)
d1 <- mr_diagnose(fit_soil)
d1$fit
d1$heteroscedasticity
d1$specification
```

Normality, heteroscedasticity and functional form are different questions.
Normal residuals do not guarantee constant variance, and constant variance does
not guarantee that the conditional mean is correctly specified.

## Example 2: intentionally heteroscedastic agronomic data

```{r}
het <- mr_example_data("agronomy_heteroskedastic")
fit_het <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  het
)
d2 <- mr_diagnose(fit_het)
d2$heteroscedasticity
```

When the mean model remains scientifically reasonable, heteroscedasticity can
be handled at the uncertainty level.

```{r}
inf_classical <- mr_inference(fit_het, "classical")
inf_hc3 <- mr_inference(fit_het, "HC3")
inf_hc4 <- mr_inference(fit_het, "HC4")
inf_classical[, c("term", "std_error", "p_value")]
inf_hc3[, c("term", "std_error", "p_value")]
inf_hc4[, c("term", "std_error", "p_value")]
```

The coefficients are unchanged because the fitted conditional mean has not
changed. Only the estimated covariance matrix and resulting uncertainty differ.
This distinction is important in agronomic field data, where variance often
changes across fertility or water gradients.

## Example 3: nonlinear irrigation response fitted linearly

```{r}
nl <- mr_example_data("irrigation_nonlinear")
fit_lin <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C +
    plant_stand_plants_m2,
  nl
)
d3 <- mr_diagnose(fit_lin)
d3$specification
```

The specification screen adds squared and cubic powers of fitted values to the
linear predictor and evaluates whether they add information. A small p-value is
not a command to use a GAM; it is evidence that the current linear functional
form deserves inspection.

## Visual diagnostics

```{r diagplot, fig.cap="Residual and influence diagnostic panels for the heteroscedastic teaching dataset."}
par(mfrow = c(2, 2))
plot(fit_het$model)
par(mfrow = c(1, 1))
```

Residual-versus-fitted plots are often more informative than a single test.
Patterns such as a funnel, curvature or systematic bands can point to different
problems and therefore different responses.

## Influence is multidimensional

```{r}
i_soil <- mr_influence(fit_soil)
head(i_soil[order(-i_soil$cooks_distance), ], 6)
```

```{r}
i_plant <- mr_influence(mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  mr_example_data("plant_growth")
))
head(i_plant[order(-i_plant$leverage), ], 6)
```

```{r}
i_het <- mr_influence(fit_het)
with(i_het, table(flag_leverage, flag_cook))
```

Leverage identifies unusual predictor combinations. Studentized residuals
identify unusual responses conditional on predictors. Cook's distance measures
the combined effect of residual size and leverage on the fitted coefficient
vector. DFFITS focuses on fitted values, while DFBETAS focus on coefficients.
No single measure should dominate interpretation.

## Sensitivity analysis rather than deletion

A defensible workflow for a flagged observation is:

1. verify the original record and measurement units;
2. determine whether the observation belongs to the target population;
3. inspect whether its predictor combination is scientifically plausible;
4. fit the prespecified model with and without the observation as a sensitivity analysis;
5. report materially changed conclusions;
6. exclude the observation only with a documented reason independent of the desired result.

The package deliberately does not implement an `auto_remove_outliers()` function.

## Cluster-robust illustration

```{r}
het$block <- rep(seq_len(18), length.out = nrow(het))
fit_cluster <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  het
)
mr_inference(fit_cluster, "cluster", cluster = "block")
```

This artificial cluster variable demonstrates syntax only. Real cluster-robust
inference must use the true independent sampling or experimental units.

## Interpretation checklist

Before modifying a model, ask whether the signal concerns the conditional mean,
residual variance, data quality, dependence structure or influential support.
The same small p-value or large Cook's distance can imply different actions under
different experimental designs.
