---
title: "Multicollinearity: VIF, condition indices and agronomic interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multicollinearity: VIF, condition indices and agronomic interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(multiRegFlow)
```

## Collinearity is not merely a threshold problem

In agronomy, correlated predictors are often expected. Soil organic matter,
CEC and clay can co-vary; N rate, N uptake and canopy N represent related
processes; plant height, leaf area and biomass can track plant vigor. The
question is not simply whether two predictors correlate, but whether the data
contain enough independent information to estimate their conditional effects.

For predictor \(X_j\), VIF is based on the coefficient of determination from
regressing \(X_j\) on the remaining predictors. Large VIF therefore indicates
that the unique variation available for estimating its slope is limited.

## Example 1: deliberately collinear N-status indicators

```{r}
d <- mr_example_data("agronomy_collinear")
fit <- mr_fit(yield_t_ha ~ ., d)
col <- mr_collinearity(fit)
col$VIF
```

```{r}
head(col$correlations, 10)
col$condition_index
```

Variance-decomposition proportions help identify which coefficient dimensions
are associated with high condition indices.

```{r}
round(col$variance_decomposition, 3)
```

A high condition index becomes more informative when two or more variables have
large variance proportions concentrated in the same high-index dimension.

## Example 2: soil fertility variables

```{r}
soil <- mr_example_data("soil_fertility")
fit_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 + soil_moisture_pct,
  soil
)
col_soil <- mr_collinearity(fit_soil)
col_soil$VIF
head(col_soil$correlations, 8)
```

Here CEC, clay and organic matter have an agronomic reason to share information.
If the goal is to estimate the effect of available P while adjusting for soil
buffering properties, dropping all correlated soil descriptors merely to reduce
VIF can change the scientific estimand.

## Example 3: plant growth indicators

```{r}
plant <- mr_example_data("plant_growth")
fit_plant <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g +
    specific_leaf_area_cm2_g,
  plant
)
mr_collinearity(fit_plant)$VIF
```

This third example demonstrates that the same diagnostics can be used when
predictors represent physiology and morphology rather than soil attributes.

## Why VIF = 5 or 10 is not a universal decision rule

Thresholds are useful for screening, but their practical meaning depends on the
scientific objective, sample size, effect size, measurement precision and the
role of the predictor. A high VIF may be acceptable for a nuisance covariate
that must remain in the model, while it may seriously limit attempts to
separate the effects of two competing treatments or nutrient indicators.

`multiRegFlow` therefore returns two flags, VIF >= 5 and VIF >= 10, but never
automatically removes a term.

## Compare OLS with ridge under collinearity

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  ridge <- mr_regularize(fit, "ridge", nfolds = 5, seed = 44)
  ridge$coefficients
}
```

Ridge does not make collinearity disappear. It changes the estimation problem by
shrinking coefficients, which can greatly improve stability and prediction. For
explanatory work, this makes ridge a useful sensitivity analysis rather than an
automatic replacement for the prespecified OLS estimand.

## Compare selection stability

```{r, eval=FALSE}
if (requireNamespace("stabs", quietly = TRUE) &&
    requireNamespace("glmnet", quietly = TRUE)) {
  stab <- mr_select(fit, "stability", cutoff = 0.75, PFER = 1)
  stab$selected
  sort(stab$selection_probabilities, decreasing = TRUE)
}
```

If two correlated indicators alternate across resamples, the instability itself
is scientifically informative. It suggests that the dataset supports a shared
signal more strongly than a unique ranking between the indicators.

## Practical decisions

When multicollinearity is substantial, consider:

- retaining scientifically mandatory variables and emphasizing their joint role;
- choosing one indicator based on measurement quality or biological priority, not p-value alone;
- creating a justified composite if several measurements represent one construct;
- using ridge for coefficient stabilization or prediction;
- using stability selection when variable-screening reproducibility is important;
- collecting data that separate the predictors more effectively in future experiments.

The final decision should follow the research question rather than an arbitrary
VIF target.
