---
title: "Variable selection, model search and stability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Variable selection, model search and stability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(multiRegFlow)
```

## Selection is a model-search operation

Variable selection can reduce a candidate model, but it does not erase the
uncertainty created by searching among many specifications. This matters most
when the selected model is subsequently interpreted as if it had been fixed in
advance. `multiRegFlow` therefore exposes selection methods but labels them as
exploratory tools unless the inferential framework explicitly accounts for the
search.

## Example 1: AIC

```{r}
soil <- mr_example_data("soil_fertility")
fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 + soil_moisture_pct +
    season_rain_mm,
  soil
)
sel_aic <- mr_select(fit, "aic")
sel_aic$selected
sel_aic$trace
```

AIC emphasizes expected predictive information loss. It can favor a larger model
than BIC, particularly with moderate sample sizes.

## Example 2: BIC

```{r}
sel_bic <- mr_select(fit, "bic")
sel_bic$selected
sel_bic$criterion
```

BIC penalizes model size more strongly as sample size increases. Neither AIC nor
BIC says that unselected variables are biologically irrelevant.

## Example 3: protect scientifically required terms

```{r}
sel_keep <- mr_select(
  fit, "aic",
  keep = c("available_P_mg_dm3", "season_rain_mm")
)
sel_keep$selected
```

A covariate can be scientifically required even if its individual test is weak.
Examples include a design variable, a known confounder, a baseline measurement
or an environmental exposure that defines the estimand.

## Best-subset exploration

```{r}
if (requireNamespace("leaps", quietly = TRUE)) {
  best <- mr_select(fit, "best_subset")
  best$variables_by_bic
  best$variables_by_adjr2
}
```

Best subsets enumerates candidate combinations efficiently. It is useful for
teaching how selection criteria can disagree. It becomes computationally
unattractive as the number of predictors grows and still does not solve
post-selection inference.

## Penalized selection

```{r}
coll <- mr_example_data("agronomy_collinear")
fit_coll <- mr_fit(yield_t_ha ~ ., coll)
if (requireNamespace("glmnet", quietly = TRUE)) {
  s_lasso_1se <- mr_select(fit_coll, "lasso", lambda = "1se", seed = 20)
  s_lasso_min <- mr_select(fit_coll, "lasso", lambda = "min", seed = 20)
  s_enet <- mr_select(fit_coll, "elastic_net", alpha = 0.5,
                      lambda = "1se", seed = 20)
  s_lasso_1se$selected
  s_lasso_min$selected
  s_enet$selected
}
```

The one-standard-error rule usually favors a simpler penalized model than the
minimum-CV-error rule. In correlated predictor groups, differences between the
two solutions are useful evidence about selection stability.

## Stability selection

```{r, eval=FALSE}
stab <- mr_select(
  fit_coll,
  method = "stability",
  cutoff = 0.75,
  PFER = 1
)
stab$selected
sort(stab$selection_probabilities, decreasing = TRUE)
```

Stability selection asks a different question: which variables continue to be
selected when the data are repeatedly perturbed through subsampling? The output
is therefore more informative than a single selected set when predictors are
numerous or correlated.

## Three levels of evidence

A useful way to interpret selection results is to distinguish:

1. **scientific necessity**: terms required by design or substantive reasoning;
2. **sample-specific evidence**: terms favored by AIC, BIC or best subsets in the current data;
3. **stability evidence**: terms repeatedly selected across resamples or penalized solutions.

A predictor supported by all three levels is different from a predictor that
appears in one stepwise solution only.

## Compare selected models out of sample

Selection criteria do not directly tell us whether predictive performance
improves.

```{r}
fit_small <- mr_fit(
  yield_t_ha ~ organic_matter_pct + available_P_mg_dm3 + season_rain_mm,
  soil,
  goal = "prediction"
)
fit_full <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 + soil_moisture_pct +
    season_rain_mm,
  soil,
  goal = "prediction"
)
mr_compare(full = fit_full, reduced = fit_small,
           v = 5, repeats = 3, seed = 20)$prediction
```

If a smaller model predicts just as well, it may be preferable for deployment.
If the goal is explanation, however, a smaller predictive model is not
necessarily a better scientific adjustment set.

## Recommended reporting

Report the initial candidate set, the selection criterion, variables forced into
the model, the final selected terms, and at least one stability or validation
analysis when selection materially affects the conclusion. Avoid describing an
algorithmically selected model as "the true model".
