---
title: "Ridge, lasso and elastic net for agronomic multiple regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ridge, lasso and elastic net for agronomic multiple regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.8)
library(multiRegFlow)
```

## Why regularize?

Ordinary least squares is unbiased under the classical linear model but can have
large sampling variance when predictors are strongly correlated or numerous
relative to the available observations. Penalized regression accepts bias in
exchange for lower variance and often better prediction.

The package uses `glmnet` as the computational backend and keeps regularization
separate from confirmatory OLS inference.

## Collinear agronomic example

```{r}
d <- mr_example_data("agronomy_collinear")
fit <- mr_fit(yield_t_ha ~ ., d, goal = "prediction")
mr_collinearity(fit)$VIF
```

The related N-status measurements create a realistic setting in which
coefficient stabilization is valuable.

## Example 1: ridge regression

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  ridge <- mr_regularize(fit, method = "ridge", nfolds = 5,
                         lambda = "1se", seed = 101)
  ridge$coefficients
  ridge$lambda_1se
}
```

Ridge retains all predictors but shrinks their coefficients. This makes it well
suited to groups of correlated measurements that all carry information.

```{r ridgecv, fig.cap="Cross-validation curve for ridge regression."}
if (exists("ridge")) plot(ridge$model)
```

## Example 2: lasso

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  lasso <- mr_regularize(fit, method = "lasso", nfolds = 5,
                         lambda = "1se", seed = 101)
  lasso$selected
  lasso$coefficients
}
```

Lasso performs shrinkage and selection. With highly correlated predictors, the
selected member of a group can change under small perturbations of the data.
That instability is not a software failure; it reflects limited information for
distinguishing predictors that encode similar signals.

## Example 3: elastic net

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  enet <- mr_regularize(fit, method = "elastic_net", alpha = 0.5,
                        nfolds = 5, lambda = "1se", seed = 101)
  enet$selected
  enet$coefficients
}
```

Elastic net often behaves more smoothly than lasso when predictors occur in
correlated groups. The mixing parameter `alpha` controls the balance between
ridge and lasso penalties.

## Lambda minimum versus one-standard-error rule

```{r}
if (requireNamespace("glmnet", quietly = TRUE)) {
  lasso_min <- mr_regularize(fit, "lasso", lambda = "min",
                             nfolds = 5, seed = 101)
  c(lambda_min = lasso_min$lambda_min,
    lambda_1se = lasso_min$lambda_1se)
}
```

`lambda.min` minimizes cross-validated loss. `lambda.1se` selects a more
regularized solution whose estimated loss is within one standard error of the
minimum. The latter is often useful when parsimony and stability matter.

## Soil example

Regularization is not limited to synthetic high-collinearity demonstrations.
A soil model with texture, organic matter and CEC can also contain partially
redundant information.

```{r}
soil <- mr_example_data("soil_fertility")
soil_fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 + season_rain_mm,
  soil,
  goal = "prediction"
)
if (requireNamespace("glmnet", quietly = TRUE)) {
  soil_ridge <- mr_regularize(soil_fit, "ridge", nfolds = 5, seed = 22)
  soil_lasso <- mr_regularize(soil_fit, "lasso", nfolds = 5, seed = 22)
  soil_lasso$selected
}
```

## Plant-science example

```{r}
plant <- mr_example_data("plant_growth")
plant_fit <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g +
    specific_leaf_area_cm2_g,
  plant,
  goal = "prediction"
)
if (requireNamespace("glmnet", quietly = TRUE)) {
  plant_enet <- mr_regularize(plant_fit, "elastic_net", alpha = 0.4,
                              nfolds = 5, seed = 23)
  plant_enet$selected
}
```

## Coefficient interpretation

Penalized coefficients should not be reported with ordinary OLS standard errors
or ordinary OLS p-values. If the scientific goal is coefficient interpretation,
use the prespecified OLS model with appropriate covariance estimation and treat
regularization as a stability or prediction analysis.

## Suggested reporting

Report the standardization behavior of the backend, penalty family, alpha,
number of CV folds, random seed, the lambda rule, selected/nonzero terms and an
out-of-sample performance summary. For correlated predictor groups, discuss the
group structure rather than over-interpreting the identity of a single selected
measurement.
