---
title: "Bootstrap uncertainty and resampling for multiple regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bootstrap uncertainty and resampling for multiple regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(multiRegFlow)
```

## Why resample?

Bootstrap procedures approximate repeated sampling by generating many datasets
from the observed data or fitted residual structure. They are useful for
coefficient uncertainty, sign stability and sensitivity analyses, but the
resampling unit must respect the experimental or sampling design.

The current bootstrap implementation targets independent observational units in
OLS models. Clustered, repeated-measures or spatial experiments require a
resampling scheme aligned with those units and should not be reduced to naive
row resampling.

## Example 1: pairs bootstrap in a soil model

```{r}
soil <- mr_example_data("soil_fertility")
fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    exchangeable_K_mg_dm3 + season_rain_mm,
  soil
)
b_pairs <- mr_bootstrap(fit, R = 250, type = "pairs", seed = 11)
b_pairs$intervals
b_pairs$sign_stability
```

Pairs bootstrap resamples entire rows. It is a natural default for independent
observational data because both response and predictor combinations are
resampled.

## Example 2: residual bootstrap

```{r}
b_resid <- mr_bootstrap(fit, R = 250, type = "residual", seed = 11)
b_resid$intervals
```

Residual bootstrap retains the observed design matrix and resamples residuals.
It is most defensible when the fitted mean is adequate and residuals are
approximately exchangeable with constant variance.

## Example 3: wild bootstrap under heteroscedasticity

```{r}
het <- mr_example_data("agronomy_heteroskedastic")
fit_het <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  het
)
mr_diagnose(fit_het)$heteroscedasticity
b_wild <- mr_bootstrap(fit_het, R = 250, type = "wild", seed = 12)
b_wild$intervals
```

The Rademacher wild bootstrap multiplies residuals by random signs. It preserves
heteroscedastic residual magnitudes more naturally than ordinary residual
resampling.

## Interval types

`multiRegFlow` provides percentile, basic and normal bootstrap summaries.

```{r}
b_pct <- mr_bootstrap(fit, R = 200, type = "pairs",
                      interval = "percentile", seed = 13)
b_basic <- mr_bootstrap(fit, R = 200, type = "pairs",
                        interval = "basic", seed = 13)
b_norm <- mr_bootstrap(fit, R = 200, type = "pairs",
                       interval = "normal", seed = 13)
```

```{r}
cbind(
  percentile = b_pct$intervals["available_P_mg_dm3", c("lower", "upper")],
  basic = b_basic$intervals["available_P_mg_dm3", c("lower", "upper")],
  normal = b_norm$intervals["available_P_mg_dm3", c("lower", "upper")]
)
```

The intervals answer similar but not identical questions. Percentile intervals
are easy to interpret but can perform poorly in some biased or skewed settings.
The basic interval reflects the bootstrap error around the original estimate.
The normal interval is compact but relies more strongly on approximate symmetry.
For advanced applications, specialized bootstrap packages may provide BCa or
design-specific resampling that is preferable.

## Sign stability

```{r}
sort(b_pairs$sign_stability)
```

A coefficient that changes sign in many resamples deserves a different
interpretation from a coefficient with a narrow distribution entirely on one
side of zero. Sign stability is descriptive evidence, not a replacement for a
confidence interval or scientific effect-size judgment.

## Bootstrap versus HC covariance

```{r}
mr_inference(fit_het, "classical")
mr_inference(fit_het, "HC3")
```

```{r}
b_wild$intervals
```

HC covariance and wild bootstrap attack the same heteroscedasticity problem from
different directions. Agreement between them can strengthen confidence that a
conclusion is not an artifact of the classical constant-variance standard error.
Disagreement should be investigated rather than hidden.

## Plant-science resampling example

```{r}
plant <- mr_example_data("plant_growth")
fit_plant <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  plant
)
b_plant <- mr_bootstrap(fit_plant, R = 200, type = "pairs", seed = 14)
b_plant$intervals
```

## Choosing the number of replicates

Small values of `R` are useful for teaching and automated examples. Final
scientific analyses should use enough replicates that reported intervals and
stability summaries do not change materially when the bootstrap is rerun with a
different seed. Values in the thousands are common, but the required number
depends on the precision needed and computational cost.

## Reproducibility

Always record the bootstrap type, resampling unit, number of replicates,
confidence level, interval type and random seed. If exclusions or transformations
are applied before resampling, document those decisions independently of the
bootstrap results.
