---
title: "Importance, effect size, marginal effects and SHAP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importance, effect size, marginal effects and SHAP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.7)
library(multiRegFlow)
```

## There is no single universal definition of importance

Researchers often ask which predictor is "most important". That question is
underspecified. A large standardized slope, a large unique contribution to
`R-squared`, a high LMG decomposition and a large average absolute SHAP value can
rank variables differently because they describe different properties of the
fitted model.

This vignette keeps these meanings separate.

## Soil-fertility model

```{r}
soil <- mr_example_data("soil_fertility")
fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  soil
)
```

## Example 1: standardized coefficients

```{r}
std <- mr_importance(fit, "standardized")
std[order(-abs(std$standardized_beta)), ]
```

A standardized coefficient expresses the conditional slope after scaling the
predictor and response by their standard deviations. It is still a coefficient
from the specified regression model and remains sensitive to collinearity.

## Example 2: partial R-squared

```{r}
pr2 <- mr_importance(fit, "partial_r2")
pr2[order(-pr2$partial_R2), ]
```

For a one-degree-of-freedom coefficient, partial R-squared describes the fraction
of residual variation associated with that coefficient after accounting for the
remaining model. It is closely related to the coefficient t statistic.

## Example 3: semi-partial R-squared

```{r}
spr2 <- mr_importance(fit, "semi_partial_r2")
spr2[order(-spr2$semi_partial_R2), ]
```

Semi-partial R-squared measures the reduction in total response variation when a
term is added to a model that already contains the other terms. Because it uses
term-level deletion, it is useful for multi-degree-of-freedom terms as well as
simple slopes.

## LMG relative importance

```{r}
if (requireNamespace("relaimpo", quietly = TRUE)) {
  lmg <- mr_importance(fit, "lmg")
  lmg$lmg
}
```

LMG averages the incremental contribution to `R-squared` over predictor orderings.
It is especially useful when predictors share explained variation. It remains a
descriptive decomposition of model fit rather than a causal ranking.

## SHAP for the additive linear model

```{r}
if (requireNamespace("kernelshap", quietly = TRUE)) {
  sh <- mr_importance(fit, "shap")
  head(sh$S)
  sort(colMeans(abs(sh$S)), decreasing = TRUE)
}
```

For an additive linear model, the package delegates to the efficient additive
SHAP route. Model-agnostic Kernel SHAP would be unnecessarily expensive here.
SHAP contributions describe model predictions relative to a baseline. A large
SHAP value does not establish that changing the predictor will cause the
predicted response to change by the same amount.

## Marginal effects: soil example

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "available_P_mg_dm3")
}
```

## Marginal effects: rainfall example

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "season_rain_mm")
}
```

## Marginal effects: plant-science example

```{r}
plant <- mr_example_data("plant_growth")
fit_plant <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  plant
)
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit_plant, variables = c("leaf_N_pct", "chlorophyll_spad"))
}
```

For a purely additive linear model, average slopes equal the corresponding
coefficients. The distinction becomes more valuable for interactions and
nonlinear models.

## Interaction example

```{r}
fit_int <- mr_fit(
  yield_t_ha ~ available_P_mg_dm3 * soil_moisture_pct +
    organic_matter_pct + season_rain_mm,
  soil
)
summary(fit_int)
```

With an interaction, the coefficient of available P no longer represents one
constant P effect across all soil moisture levels. Marginal-effect tools can
summarize the average slope or evaluate the slope at meaningful moisture values.

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit_int, variables = "available_P_mg_dm3")
}
```

## Compare importance under collinearity

```{r}
coll <- mr_example_data("agronomy_collinear")
fit_coll <- mr_fit(yield_t_ha ~ ., coll)
mr_importance(fit_coll, "standardized")
mr_importance(fit_coll, "semi_partial_r2")
```

Under strong collinearity, unique-contribution measures can become small even
when a group of predictors jointly carries substantial information. This is one
reason not to interpret importance rankings without the collinearity analysis.

## Reporting recommendations

State exactly which definition of importance is used. Report uncertainty where
available. Avoid phrases such as "X caused most of the yield variation" when the
analysis only supports a model-based descriptive ranking. When predictors have
different units or belong to correlated groups, show more than one importance
view and explain why rankings differ.
