---
title: "Robust, quantile, GAM and kernel alternatives"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust, quantile, GAM and kernel alternatives}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.7)
library(multiRegFlow)
```

## Alternatives should answer a diagnosed problem

A flexible model is not automatically better than OLS. Different alternatives
address different departures:

- robust regression reduces the influence of unusual observations;
- quantile regression changes the target from the conditional mean to a conditional quantile;
- GAM relaxes linear functional forms while retaining an additive structure;
- kernel regression provides a more nonparametric conditional-mean estimate.

The first task is therefore to identify why an alternative is scientifically
useful.

## Robust regression: soil example

```{r}
soil <- mr_example_data("soil_fertility")
ols_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    season_rain_mm,
  soil
)
rob_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + available_P_mg_dm3 +
    season_rain_mm,
  soil,
  engine = "robust"
)
cbind(OLS = coef(ols_soil), Robust = coef(rob_soil))
```

Robust regression is best interpreted as a sensitivity analysis when unusual
observations may disproportionately affect OLS. It should not be used to hide
valid observations simply because they disagree with the dominant trend.

## Robust regression: plant example

```{r}
plant <- mr_example_data("plant_growth")
rob_plant <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm,
  plant,
  engine = "robust"
)
summary(rob_plant)
```

## Quantile regression: low, median and high yield

```{r}
het <- mr_example_data("agronomy_heteroskedastic")
base_het <- mr_fit(
  grain_yield_t_ha ~ N_rate_kg_ha + P_rate_kg_ha + soil_water_m3_m3,
  het
)
if (requireNamespace("quantreg", quietly = TRUE)) {
  q25 <- mr_quantile(base_het, tau = 0.25)
  q50 <- mr_quantile(base_het, tau = 0.50)
  q75 <- mr_quantile(base_het, tau = 0.75)
  coef(q25$model)
  coef(q50$model)
  coef(q75$model)
}
```

The three calls demonstrate that `mr_quantile()` can be used for distinct
conditional response levels. A single multi-quantile fit is also available.

```{r}
if (requireNamespace("quantreg", quietly = TRUE)) {
  q_all <- mr_quantile(base_het, tau = c(0.1, 0.25, 0.5, 0.75, 0.9))
  summary(q_all$model)
}
```

Quantile-specific slopes can be agronomically relevant when nutrient or water
responses differ between stressed and high-performing environments.

## Detect nonlinear irrigation response

```{r}
nl <- mr_example_data("irrigation_nonlinear")
lin <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C,
  nl,
  goal = "prediction"
)
mr_diagnose(lin)$specification
```

## GAM: interpretable smooth effects

```{r gamfit}
gam1 <- mr_fit(
  yield_t_ha ~ s(irrigation_mm) + s(N_rate_kg_ha) + s(mean_temp_C),
  nl,
  engine = "gam",
  goal = "prediction"
)
summary(gam1)
```

```{r gamplot, fig.cap="Estimated smooth response functions in the irrigation GAM."}
plot(gam1$model, pages = 1, residuals = TRUE)
```

A second GAM can retain one linear agronomic term and smooth only the variables
with plausible nonlinear response.

```{r}
gam2 <- mr_fit(
  yield_t_ha ~ s(irrigation_mm) + N_rate_kg_ha + s(mean_temp_C) +
    plant_stand_plants_m2,
  nl,
  engine = "gam",
  goal = "prediction"
)
summary(gam2)
```

## Kernel regression

Kernel regression is useful when the functional form is difficult to specify.
Bandwidth selection is part of the model and can be computationally expensive.
The following three examples are kept unevaluated during vignette building to
avoid long package-installation times; the package validation script executes
them separately.

```{r, eval=FALSE}
ker1 <- mr_kernel(lin)
ker1$bandwidth
```

```{r, eval=FALSE}
lin_two <- mr_fit(yield_t_ha ~ irrigation_mm + N_rate_kg_ha, nl)
ker2 <- mr_kernel(lin_two)
ker2$model
```

```{r, eval=FALSE}
lin_temp <- mr_fit(yield_t_ha ~ irrigation_mm + mean_temp_C, nl)
ker3 <- mr_kernel(lin_temp, bwmethod = "cv.aic")
ker3$bandwidth
```

## Compare OLS, GAM and kernel out of sample

```{r, eval=FALSE}
ker_fit <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C,
  nl,
  engine = "kernel",
  goal = "prediction"
)
mr_compare(linear = lin, gam = gam1, kernel = ker_fit,
           v = 5, repeats = 3, seed = 91)$prediction
```

The comparison must use out-of-fold predictions. A flexible model can always
fit training data more closely, but that is not enough to demonstrate improved
prediction.

## Choosing among alternatives

Use robust regression when the main concern is sensitivity to influential
observations. Use quantile regression when the scientific target is a conditional
quantile. Use GAM when smooth nonlinear functional forms are plausible and
interpretability remains important. Use kernel regression when a less structured
conditional-mean estimate is justified and sample size supports it.

A single dataset may motivate more than one sensitivity analysis, but every
additional model should correspond to an explicit scientific or diagnostic
question.
