---
title: "Out-of-sample validation and fair model comparison"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Out-of-sample validation and fair model comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(multiRegFlow)
```

## Training fit is not predictive validation

A model evaluated on the same observations used to estimate its coefficients
receives an optimistic assessment of predictive performance. `mr_validate()`
uses repeated cross-validation so every reported prediction is generated by a
model that did not use that observation for fitting.

The package reports RMSE, MAE, out-of-fold R-squared, calibration intercept and
calibration slope.

## Example 1: soil yield prediction

```{r}
soil <- mr_example_data("soil_fertility")
soil_full <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  soil,
  goal = "prediction"
)
cv_soil <- mr_validate(soil_full, v = 5, repeats = 4, seed = 50)
cv_soil$summary
```

## Example 2: plant biomass prediction

```{r}
plant <- mr_example_data("plant_growth")
plant_fit <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  plant,
  goal = "prediction"
)
cv_plant <- mr_validate(plant_fit, v = 5, repeats = 4, seed = 51)
cv_plant$summary
```

## Example 3: nonlinear GAM validation

```{r}
nl <- mr_example_data("irrigation_nonlinear")
gam_fit <- mr_fit(
  yield_t_ha ~ s(irrigation_mm) + s(N_rate_kg_ha) + s(mean_temp_C),
  nl,
  engine = "gam",
  goal = "prediction"
)
cv_gam <- mr_validate(gam_fit, v = 5, repeats = 3, seed = 52)
cv_gam$summary
```

The same interface can validate robust regression and a single-quantile
regression. Kernel models are also supported, although repeated bandwidth
selection can be computationally expensive.

## Observed versus out-of-fold predictions

```{r predplot, fig.cap="Observed and out-of-fold predicted soil yield from repeated cross-validation."}
p <- cv_soil$predictions
plot(p$observed, p$predicted,
     xlab = "Observed yield (t/ha)",
     ylab = "Out-of-fold predicted yield (t/ha)")
abline(0, 1, lty = 2)
```

The plot includes predictions from multiple repetitions. It visualizes both
bias and dispersion rather than compressing predictive behavior into one number.

## Calibration

A calibration intercept near zero indicates little systematic shift in the mean
prediction. A calibration slope near one indicates that prediction spread is
appropriately scaled. A slope below one often indicates predictions that are too
extreme, while a slope above one can indicate predictions that are too narrow.
Calibration is a diagnostic, not a universal acceptance test.

## Model comparison 1: full versus reduced soil model

```{r}
soil_small <- mr_fit(
  yield_t_ha ~ organic_matter_pct + available_P_mg_dm3 + season_rain_mm,
  soil,
  goal = "prediction"
)
cmp1 <- mr_compare(full = soil_full, reduced = soil_small,
                   v = 5, repeats = 3, seed = 60)
cmp1$inference
cmp1$prediction
```

AIC/BIC and out-of-fold metrics appear in separate tables because they are not
interchangeable.

## Model comparison 2: linear versus GAM irrigation response

```{r}
lin_fit <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C,
  nl,
  goal = "prediction"
)
cmp2 <- mr_compare(linear = lin_fit, gam = gam_fit,
                   v = 5, repeats = 3, seed = 61)
cmp2$prediction
```

## Model comparison 3: OLS versus robust regression

```{r}
coll <- mr_example_data("agronomy_collinear")
ols <- mr_fit(
  yield_t_ha ~ N_rate_kg_ha + season_rain_mm + soil_mineral_N_mg_kg,
  coll,
  goal = "prediction"
)
rob <- mr_fit(
  yield_t_ha ~ N_rate_kg_ha + season_rain_mm + soil_mineral_N_mg_kg,
  coll,
  engine = "robust",
  goal = "prediction"
)
cmp3 <- mr_compare(OLS = ols, robust = rob,
                   v = 5, repeats = 3, seed = 62)
cmp3$prediction
```

Robust regression can improve prediction when extreme observations distort OLS,
but it need not outperform OLS in every dataset. The comparison should be made
with the same folds, response and predictor information.

## What not to compare directly

Do not compare training `R-squared` from one model with cross-validated RMSE from
another. Do not choose a model solely because its AIC is lower if the stated goal
is prediction. Conversely, a slightly smaller predictive RMSE does not justify
removing a scientifically required adjustment variable from an explanatory
model.

## Repeated CV and uncertainty

`mr_validate()` reports metrics for each repetition and summarizes their mean and
standard deviation. The repetition-level variation is not a full confidence
interval for generalization performance, but it provides a useful measure of
sensitivity to fold assignment.

```{r}
cv_soil$metrics
```

For high-stakes prediction, nested cross-validation or an untouched external
test set may be preferable, particularly when hyperparameters and feature sets
are tuned extensively.
