---
title: "Integrated soil-fertility case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated soil-fertility case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.7)
library(multiRegFlow)
```

## Scientific setting

This vignette treats the synthetic `soil_fertility` data as a teaching analogue
of an observational agronomic dataset. The response is crop yield in t/ha and
the candidate predictors represent soil pH, organic matter, clay, CEC,
available P, exchangeable K, soil moisture and seasonal rainfall.

The data are synthetic. The interpretation below illustrates workflow, not a
claim about a real site or crop.

```{r}
d <- mr_example_data("soil_fertility")
summary(d)
```

## Question 1: conditional nutrient and soil associations

```{r}
fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  d,
  goal = "explanation"
)
summary(fit)
```

The coefficient table should be read together with uncertainty and model
adequacy. A small p-value is not the sole criterion for retaining a soil
variable, particularly when it represents an adjustment factor or a process that
is scientifically required.

## Diagnostics

```{r}
dg <- mr_diagnose(fit)
dg$fit
dg$heteroscedasticity
dg$specification
```

```{r}
inf <- mr_influence(fit)
head(inf[order(-inf$cooks_distance), ], 8)
```

The influential-observation table should trigger data verification and
sensitivity analysis, not automatic deletion.

## Collinearity among soil descriptors

```{r}
co <- mr_collinearity(fit)
co$VIF
head(co$correlations, 10)
```

CEC, clay and organic matter can share information because cation exchange
capacity is related to mineral and organic colloids. A high VIF in a real soil
dataset could therefore reflect a genuine soil process rather than a defective
variable.

## Coefficient inference

```{r}
mr_inference(fit, "classical")
```

```{r}
mr_inference(fit, "HC3")
```

```{r}
mr_inference(fit, "HC4")
```

The three calls illustrate classical and two heteroscedasticity-consistent
covariance estimators. Robust covariance does not change the OLS coefficient;
it changes the estimated uncertainty.

## Importance from three perspectives

```{r}
mr_importance(fit, "standardized")
mr_importance(fit, "partial_r2")
mr_importance(fit, "semi_partial_r2")
```

These measures can rank variables differently. The reporting language should
name the measure rather than stating that a variable is simply "the most
important".

## Marginal effects

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "available_P_mg_dm3")
  mr_effects(fit, variables = "organic_matter_pct")
  mr_effects(fit, variables = "season_rain_mm")
}
```

For this additive linear specification, these average slopes reproduce the
corresponding conditional slopes. The same interface becomes more informative
with interactions and nonlinear terms.

## Selection as sensitivity analysis

```{r}
aic <- mr_select(fit, "aic")
bic <- mr_select(fit, "bic")
aic$selected
bic$selected
```

```{r}
if (requireNamespace("leaps", quietly = TRUE)) {
  best <- mr_select(fit, "best_subset")
  best$variables_by_bic
}
```

The selected set should be compared with the scientifically prespecified model.
A variable that disappears under BIC is not thereby biologically irrelevant.

## Bootstrap stability

```{r}
b <- mr_bootstrap(fit, R = 250, type = "pairs", seed = 71)
b$intervals
sort(b$sign_stability)
```

This table helps identify coefficients whose direction is stable across
resampled datasets and those whose uncertainty is substantial.

## Prediction-oriented refit

If the objective changes to prediction, the same predictor set should be judged
out of sample.

```{r}
pred_fit <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct + CEC_cmolc_dm3 +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  d,
  goal = "prediction"
)
mr_validate(pred_fit, v = 5, repeats = 4, seed = 72)$summary
```

## A reduced predictive model

```{r}
reduced <- mr_fit(
  yield_t_ha ~ organic_matter_pct + available_P_mg_dm3 + season_rain_mm,
  d,
  goal = "prediction"
)
mr_compare(full = pred_fit, reduced = reduced,
           v = 5, repeats = 3, seed = 73)$prediction
```

A reduced model that predicts similarly may be attractive for operational soil
sampling because fewer measurements are needed. That operational argument is
different from claiming that excluded soil variables have no scientific role.

## Reporting template

A concise soil-regression results section should report the response and units,
model formula, sample size, coefficient estimates with uncertainty, diagnostics,
collinearity evidence, influential-observation sensitivity, any selection or
shrinkage analysis, and out-of-sample performance if prediction is claimed.

```{r}
cat(substr(mr_report(fit), 1, 1200))
```
