---
title: "Integrated plant-science and crop-growth case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated plant-science and crop-growth case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.7)
library(multiRegFlow)
```

## Scientific setting

The synthetic `plant_growth` dataset links biomass to morphological and
physiological indicators: leaf area, SPAD chlorophyll index, leaf N,
stomatal conductance, plant height, root mass and specific leaf area. The
workflow illustrates how multiple regression can combine measurements from
plant nutrition, physiology and growth analysis.

```{r}
d <- mr_example_data("plant_growth")
head(d)
```

## Explanatory reference model

```{r}
fit <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g +
    specific_leaf_area_cm2_g,
  d,
  goal = "explanation"
)
summary(fit)
```

A slope represents a conditional association. For example, the leaf-N slope is
the expected biomass difference associated with a one-unit difference in leaf N
when the other included physiological and morphological variables are held
fixed. This conditional quantity can differ from the marginal correlation
between leaf N and biomass.

## Model adequacy

```{r}
mr_diagnose(fit)$fit
mr_diagnose(fit)$heteroscedasticity
mr_diagnose(fit)$specification
```

The three calls emphasize that fit, residual variance and functional form are
separate aspects of adequacy.

## Influence in multivariate plant measurements

```{r}
infl <- mr_influence(fit)
head(infl[order(-infl$cooks_distance), ], 8)
```

```{r}
subset(infl, flag_leverage | flag_cook | flag_dffits | flag_dfbeta)
```

A plant with an unusual combination of leaf area, conductance and root mass may
have high leverage even if its biomass is not an obvious univariate outlier.
This is why influence diagnostics are multivariate.

## Collinearity among plant traits

```{r}
co <- mr_collinearity(fit)
co$VIF
head(co$correlations, 10)
```

Growth traits often share biological information. High collinearity can make it
difficult to separate a unique plant-height effect from a leaf-area effect even
when the two traits jointly predict biomass well.

## Importance views

```{r}
mr_importance(fit, "standardized")
mr_importance(fit, "partial_r2")
mr_importance(fit, "semi_partial_r2")
```

Use the names of the measures in reporting. A standardized slope is not the same
quantity as unique explained variance.

## Marginal effects

```{r}
if (requireNamespace("marginaleffects", quietly = TRUE)) {
  mr_effects(fit, variables = "leaf_N_pct")
  mr_effects(fit, variables = "chlorophyll_spad")
  mr_effects(fit, variables = "stomatal_conductance")
}
```

## Robust-regression sensitivity

```{r}
rob <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g +
    specific_leaf_area_cm2_g,
  d,
  engine = "robust"
)
cbind(OLS = coef(fit), Robust = coef(rob))
```

Large changes suggest that a small number of observations may be influential.
Agreement does not prove the absence of influence, but it provides useful
sensitivity evidence.

## Regularization for prediction

```{r}
pred <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g +
    specific_leaf_area_cm2_g,
  d,
  goal = "prediction"
)
if (requireNamespace("glmnet", quietly = TRUE)) {
  ridge <- mr_regularize(pred, "ridge", nfolds = 5, seed = 81)
  lasso <- mr_regularize(pred, "lasso", nfolds = 5, seed = 81)
  enet <- mr_regularize(pred, "elastic_net", alpha = 0.5,
                        nfolds = 5, seed = 81)
  lasso$selected
  enet$selected
}
```

## Out-of-fold predictive performance

```{r}
cv <- mr_validate(pred, v = 5, repeats = 4, seed = 82)
cv$summary
```

The model should be described as predictive only on the basis of these
out-of-sample predictions or an independent test dataset, not the training
`R-squared`.

## Reduced sensor set

Suppose chlorophyll, leaf N, plant height and leaf area are relatively easy to
collect, while stomatal conductance and root mass are expensive or destructive.
A smaller predictive model can be evaluated as an operational alternative.

```{r}
reduced <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct + plant_height_cm,
  d,
  goal = "prediction"
)
mr_compare(full = pred, reduced = reduced,
           v = 5, repeats = 3, seed = 83)$prediction
```

A small loss in accuracy may be acceptable if field measurement becomes much
simpler. That trade-off should be described explicitly rather than disguised as
a purely statistical model-selection decision.

## Bootstrap coefficient stability

```{r}
b <- mr_bootstrap(fit, R = 220, type = "pairs", seed = 84)
b$intervals
sort(b$sign_stability)
```

## Integrated recommendation

```{r}
mr_recommend(fit)
```

The output is intended to support a methods discussion: whether traits are
redundant, whether robust regression changes conclusions, whether the mean
functional form is adequate and whether predictive claims have been validated.

## Reporting language

A plant-science manuscript should identify the experimental unit, growth stage,
measurement method and units for each predictor. Statistical results should
emphasize effect estimates and uncertainty, avoid causal language unsupported by
design, and distinguish explanatory models from sensor-based predictive models.
