---
title: "Reporting, audit trails and reproducible regression workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reporting, audit trails and reproducible regression workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.6)
library(multiRegFlow)
```

## Why an audit trail matters

A multiple-regression result is not just a final coefficient table. Reproducible
analysis requires a record of the scientific goal, candidate variables, model
formula, diagnostics, treatment of heteroscedasticity, influence assessment,
selection or regularization decisions, resampling settings and predictive
validation.

`mr_report()` creates a compact Markdown record from an OLS reference model. It
is designed to accompany, not replace, scientific interpretation.

## Example 1: explanatory soil report

```{r}
soil <- mr_example_data("soil_fertility")
fit_soil <- mr_fit(
  yield_t_ha ~ soil_pH + organic_matter_pct + clay_pct +
    available_P_mg_dm3 + exchangeable_K_mg_dm3 +
    soil_moisture_pct + season_rain_mm,
  soil,
  goal = "explanation"
)
rep1 <- mr_report(fit_soil)
cat(substr(rep1, 1, 1600))
```

When heteroscedasticity is detected and `sandwich` is available, the default
`vcov = "auto"` changes the reported coefficient uncertainty to HC3 while
leaving the OLS coefficients unchanged.

## Example 2: explicitly classical report

```{r}
rep2 <- mr_report(fit_soil, vcov = "classical")
cat(substr(rep2, 1, 800))
```

This is useful when the analyst needs to show exactly which covariance estimator
was used rather than rely on the automatic choice.

## Example 3: predictive report with cross-validation

```{r}
plant <- mr_example_data("plant_growth")
fit_pred <- mr_fit(
  biomass_g ~ leaf_area_cm2 + chlorophyll_spad + leaf_N_pct +
    stomatal_conductance + plant_height_cm + root_mass_g,
  plant,
  goal = "prediction"
)
rep3 <- mr_report(fit_pred, validation = TRUE,
                  v = 5, repeats = 3, seed = 90)
cat(substr(rep3, 1, 1200))
```

## Writing the report to disk

```{r}
tmp1 <- tempfile(fileext = ".md")
mr_report(fit_soil, file = tmp1)
file.exists(tmp1)
```

```{r}
tmp2 <- tempfile(fileext = ".md")
mr_report(fit_pred, file = tmp2, validation = TRUE,
          v = 5, repeats = 2, seed = 91)
file.info(tmp2)$size
```

A project workflow can store these reports next to scripts, frozen data and
session information.

## Recommendation examples

The recommendation engine is most useful when compared across different
statistical problems.

```{r}
mr_recommend(fit_soil)
```

```{r}
coll <- mr_example_data("agronomy_collinear")
fit_coll <- mr_fit(yield_t_ha ~ ., coll)
mr_recommend(fit_coll)
```

```{r}
nl <- mr_example_data("irrigation_nonlinear")
fit_nl <- mr_fit(
  yield_t_ha ~ irrigation_mm + N_rate_kg_ha + mean_temp_C,
  nl
)
mr_recommend(fit_nl)
```

The outputs differ because the problems differ. Collinearity motivates
stability analysis, a functional-form signal motivates nonlinear sensitivity,
and an explanatory goal discourages treating automated selection as confirmatory.

## Capability audit

```{r}
cap <- mr_capabilities()
cap
```

```{r}
subset(cap, !available)
```

```{r}
table(cap$backend)
```

This information is useful when a report depends on optional backends that may
not be installed on another computer.

## Minimum reproducibility record

For an analysis intended for publication or project review, save at least:

- the original or analysis-ready data with provenance;
- the exact model formula and declared goal;
- package version and `sessionInfo()`;
- random seeds for bootstrap and validation;
- bootstrap type and number of replicates;
- covariance estimator used for inference;
- selection criterion, lambda rule or stability-selection settings;
- all exclusions and transformations with justification;
- out-of-fold metrics for predictive claims;
- scripts that regenerate tables and figures.

## Suggested Methods wording

A transparent Methods description should state that multiple linear regression
was fitted as a reference model, specify the candidate predictors, describe
residual and influence diagnostics, explain how multicollinearity was evaluated,
identify any robust covariance or bootstrap analysis, and state whether variable
selection or regularization was exploratory. Predictive analyses should describe
cross-validation design and metrics.

## Suggested Results wording

Results should focus on estimates, uncertainty, effect sizes and patterns that
address the scientific question. Avoid repeating every diagnostic statistic.
Report material sensitivity to influential observations, covariance estimators,
selection procedures or flexible alternatives when those analyses change the
scientific conclusion.

## What the automated report intentionally does not do

`mr_report()` does not infer the experimental design, decide whether a predictor
is a confounder, convert associations into causal effects, determine biological
relevance from p-values, or decide whether an influential observation is invalid.
Those decisions remain the analyst's scientific responsibility.

## Final integrated audit

```{r}
list(
  diagnostics = mr_diagnose(fit_soil),
  collinearity = mr_collinearity(fit_soil)$VIF,
  importance = mr_importance(fit_soil, "semi_partial_r2"),
  recommendation = mr_recommend(fit_soil)
)
```

This final object demonstrates the package philosophy: preserve separate evidence
streams and combine them through interpretation rather than collapsing them into
one automatic score.
